---
title: "Something's Fishy"
output:
  html_document:
    df_print: paged
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r message=F, warning=F, echo=F}
library(dplyr)
library(usmap)
library(ggplot2)
```

From [538 Riddler: Can You Find The Fish In State Names?](https://fivethirtyeight.com/features/somethings-fishy-in-the-state-of-the-riddler/)

> From Mark Bradwin comes a fishy puzzle about state names:
> 
> Ohio is the only state whose name doesn’t share any letters with the 
> word “mackerel.” It’s strange, but it’s true.
> 
> But that isn’t the only pairing of a state and a word you can say that 
> about — it’s not even the only fish! Kentucky has “goldfish” to itself,
> Montana has “jellyfish” and Delaware has “monkfish,” just to name a few.
> 
> What is the longest “mackerel?” That is, what is the longest word that 
> doesn’t share any letters with exactly one state? (If multiple “mackerels” 
> are tied for being the longest, can you find them all?)
> 
> Extra credit: Which state has the most “mackerels?” That is, which state 
> has the most words for which it is the only state without any letters 
> in common with those words?
> 
> (For both the Riddler and the extra credit, please refer to Friend of 
> the Riddler™ Peter Norvig’s [word list](https://norvig.com/ngrams/word.list).)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Answers first!
1. The longest mackerels is... well, there are two, at 23 letters: "counterproductivenesses" for Alabama, "hydrochlorofluorocarbon" for Mississippi.
2. Extra credit: Ohio has the most mackerels: 11,342 of them.

## Showing the work
### Data capturing, processing, caching
First things first: I'll grab Peter Norvig's word list and derive some useful information from it. I don't want to hammer norvig.com too much during development, so I'll cache the word list locally.

```{r}
if (!file.exists("words.csv")) {
  words <- read.table("https://norvig.com/ngrams/word.list", stringsAsFactors = F)
  names(words) <- c("word")
  words$length <- nchar(words$word)
  # For an optimization, I could throw away words shorter than "mackerel", but... 
  # I need them for the extra credit question. 

  # The words are in alphabetical order; sort by length instead.
  words <- words[order(words$length, decreasing=T),]
  write.table(words, "words.csv")
}
words <- read.table("words.csv", stringsAsFactors=F)

str(words)
```

Along with all the words, I need all the state names. I made myself a list of the 50 states, all lower case.

```{r}
states <- read.table("states.csv", stringsAsFactors=F, sep="\n")
names(states) <- c("state")
str(states)
```

### Encoding the Codes
I'll be asking this question: "does word W have any letters in common with state S?". For example, "mackerel" has no letters in common with "ohio", but it has letters in common with maine, michigan, montana... and every other state. I'll take advantage of the fact that repeated letters, and letter order don't matter - that is, "does 'mackerel' have any letters in common with 'ohio'?" and "does 'aceklmr' have any letters in common with 'hio'?" are really the same question. I'll encode each word and state as a vector of 26 bits, with '1' meaning "letter is present". Then I'll bitwise-AND a word and a state code; iff there are no letters in common, the result is 0. If the compiler is smart, these bitwise operations should be efficient.

Here's a utility function to encode a string of letters.

```{r}
# encode each letter as a binary number with only one bit set.
# fortunately there are fewer than 33 letters, so all the encodings will fit in a 32-bit word.
bitmasks <- sapply(1:26, function(x) as.integer(2**(x-1)))
# each word is encoded as a binary number with a bit set to 1 only if a letter is present.
# examples:
#  "a" is encoded as 1
#  "ab" is encoded as 3
# "zebra" is encoded as 0x2020013
# note: encodings are not unique! That is, many words can share the same encoding, like "stars" and "rats"
encode <- function(word) {
  # list of characters
  chs <- strsplit(word, "")[[1]]
  # eliminate dups, so we can sum instead of OR
  chs <- unique(chs)
  # convert to letter position
  indices <- match(chs, letters)
  # whitespace yields NA; drop those
  indices <- indices[!is.na(indices)]
  # index to onehot bit mask
  onehots <- bitmasks[indices]
  # they're unique, so OR them by adding
  code <- sum(onehots)
  # a sum of integers might exceed integer range - but I know it won't. Cast to int for efficiency.
  return(as.integer(code))
} 
```

Now I'll augment my data sets with encodings for each word and state. With around 260,000 words in the list, I worried that this step would be very time consuming, but it only takes a few seconds.

```{r}
words$code <- sapply(words$word, encode)
states$code <- sapply(states$state, encode)
```

Here's a function that takes a candidate word code, and returns the state name for the word, _if_ the word is a mackerel. Otherwise it returns NA.

```{r}
is_mackerel <- function(code) {
  noshares <- states[bitwAnd(code, states$code) == 0,]
  if (nrow(noshares) == 1) {
    return(noshares[1, "state"])
  }
  return(NA)
}
```

### Sanity Checks 
Is "mackerel" a mackerel? If so, for which state?

```{r}
test_mackerel <- function(word, code) {
  state_or_NA <- is_mackerel(code)
  if (is.na(state_or_NA)) {
    print(sprintf("'%s' is not a mackerel!", word), quote=F)
  } else {
    print(sprintf("'%s' is a mackerel for %s", word, state_or_NA), quote=F)
  }
}
word <- "mackerel"
test_mackerel(word, encode(word))
```

And how about a non-mackerel? I'm not sure if there are very many words that have letters in common with _all_ states, but it's pretty easy to find a word with no letters in common with more than one state. How about "the"?
```{r}
word <- "the"
test_mackerel(word, encode(word))
```

I have three other examples, so I'll try them too:
```{r}
word <- "goldfish"
test_mackerel(word, encode(word))

word <- "jellyfish"
test_mackerel(word, encode(word))

word <- "monkfish"
test_mackerel(word, encode(word))

```

Sanity achieved!

### Find Some Longest Mackerels
I'll start with the longest word(s), and work my way down the list, stopping at the first mackerels. "jellyfish" is the longest mackerel I already know, so I can stop when I reach words of that length.

```{r}
for (len in max(words$length):nchar("jellyfish")) {
  longwords <- words[words$length == len,]
  if (nrow(longwords) > 0) {
    longwords$mackerel_state <- sapply(longwords$code, is_mackerel)
    mackerels <- longwords[!is.na(longwords$mackerel_state), ]
    if (nrow(mackerels) > 0) {
      print(mackerels)
      break;
    }
  }
}
```
That's it, two 23-letter words tie for first place.

### Extra credit
Which state has the most mackerels? To answer this I'm going to find all the mackerels among the 260,000 words, keeping track of each mackerel's state; then I'll find the state that appears the most times. Running is_mackerel on each word does take a while - an optimization would take advantage of the fact that many words share the same code, and only process unique codes.

```{r}
words$mackerel_state <- sapply(words$code, is_mackerel)
mackerels <- words[!is.na(words$mackerel_state), ]
```

Now I just have to add up the number of times each state is a mackerel state, for all the words. 
```{r}
mackerel_count <- mackerels %>% group_by(mackerel_state) %>% summarize(mackerels=n())
mackerel_count <- mackerel_count[order(mackerel_count$mackerels, decreasing=T),]
names(mackerel_count) <- c("state", "mackerels")
head(mackerel_count, n=10)
```

Looks like Ohio is the mackereliest; with only 3 unique letters, maybe it's not surprising.

### A Few Misguided Plots
There's quite a range of mackerel counts among the states, so I'll put them on a log scale.

```{r}
mackerel_count$fstate <- factor(mackerel_count$state, levels=mackerel_count$state[order(mackerel_count$mackerels, decreasing=T)])
ggplot(mackerel_count, aes(x=fstate, y=log10(mackerels))) +
  ggtitle("Ohio Is The Mackereliest") +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

This seems like a terrible application for a map graph (choropleth), since the data has nothing to do with geographical position. Let's do it! I'm giving Ohio a tasteful red border, to help it stand out from the crowd.

```{r}
map <- usmap::us_map()
# change the 'full' column to lowercase 'state'
map$state <- tolower(map$full)
map <- subset(map, select=-full)
map <- subset(map, select=-abbr)
map <- left_join(map, mackerel_count, by="state")
# states with no mackerels were not present in mackerel_count, so
# are NA in the joined dataframe. Set those NAs to 1 (not 0, because
# I'm going to plot on a log scale).
map$mackerels <- ifelse(is.na(map$mackerels), 1, map$mackerels)
ggplot() + 
  geom_polygon(data=map, aes(x=x, y=y, group=group, fill=log10(mackerels))) +
  geom_polygon(data=map[map$state == "ohio",], aes(x=x, y=y, group=group, fill=log10(mackerels)), color="red") +
  ggtitle("Ohio Is The Mackereliest") +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank()
  )
```

